---
title: "Spirals"
author: "Susan Vanderplas"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Logarithmic Spiral

The logarithmic spiral has a polar equation of $$r = a e^{b\theta}$$ where $r$ is distance from the origin, $\theta$ is the angle from the x-axis, and $a$ and $b$ are constants. [Source](https://mathworld.wolfram.com/LogarithmicSpiral.html)

We can write the logarithmic spiral equation in Cartesian coordinates as: $$x = r \cos\theta = a \cos\theta e^{b\theta}\\y = r \sin\theta = a \sin\theta e^{b\theta}$$

We can implement this spiral in R by first defining it in polar coordinates and then converting the polar coordinates into a sequence of cartesian points that should be connected by line segments.

```{r define-spiral-polar}
# Define the angle of the spiral (polar coords)
# go around two full times (2*pi = one revolution)
theta <- seq(0, 4*pi, .01)
# Define the distance from the origin of the spiral
# Needs to have the same length as theta
r <- seq(0, 5, length.out = length(theta))
```


```{r convert-to-cartesian}
# Now define x and y in cartesian coordinates
x <- r * cos(theta)
y <- r * sin(theta)
```

```{r plot-the-spiral}
plot(x, y, type = "l")
```
